Finally Tagless, Partially Evaluated

### 2008-10-31 tagless/tagless-final.tex
add ref to Consel
just a few minor edits

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator as follows.
In the phase tags |S0| and~|D0|, the digit zero indicates
our initial attempt.
\begin{code}
type ('c,'dv) repr = S0 of ('c,'dv) R.repr | D0 of ('c,'dv) C.repr
\end{code}
To extract a dynamic term from this type, we create the function
\begin{code}
let abstrI0 (e : ('c,int) repr) : ('c,int) C.repr =
  match e with S0 e -> C.int e | D0 e -> e
\end{code}
and the similar function |abstrB0| for dynamic boolean terms. Here, 
|C.int| is used to convert a static term (of type |('c,int) R.repr|, 
which is just |int|) to a dynamic term. We can now define the following
components required by the |Semantics| signature:
\begin{code}
let int  (x:int)  = S0 (R.int x)
let bool (x:bool) = S0 (R.bool x)
let add e1 e2 = match (e1,e2) with
  | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
  | _ -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|.

@@

Our partial evaluator owes much to
\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducers for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen,Mogensen-SelfApplicable} also
build static and dynamic terms in tandem.  However, they
build a static term for every object term, even a bound variable,
so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remain untyped.  In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.

@@

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |static|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |static|, so again we need to change the type arguments to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.
The verbosity of these types is the only difficulty in defining a
new signature to replace |Symantics| that the output of |CPST| matches.

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.
The techniques we use to form such a family of interpreters
find their seeds in \citearound{'s \emph{language
triplets}}\citet{Holst-AMIX}, though in an untyped setting.
\Citet{JonesNielson:94:AbstractInterpretation} also
prefigured this separation when they decomposed a denotational definition of an
untyped object language into a core semantics (which we call abstract syntax)
and multiple interpretations.

>>>>

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator as follows.
In the phase tags |S0| and~|D0|, the digit zero indicates
our initial attempt.
\begin{code}
type ('c,'dv) repr = S0 of ('c,'dv) R.repr | D0 of ('c,'dv) C.repr
\end{code}
To extract a dynamic term from this type, we create the function
\begin{code}
let abstrI0 (e : ('c,int) repr) : ('c,int) C.repr =
  match e with S0 e -> C.int e | D0 e -> e
\end{code}
and a similar function |abstrB0| for dynamic boolean terms. Here, 
|C.int| is used to convert a static term (of type |('c,int) R.repr|, 
which is just |int|) to a dynamic term. We can now define the following
components required by the |Symantics| signature:
\begin{code}
let int  (x:int)  = S0 (R.int x)
let bool (x:bool) = S0 (R.bool x)
let add e1 e2 = match (e1,e2) with
  | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
  | _ -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|.

@@

Our partial evaluator owes much to
\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducers for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen,Mogensen-SelfApplicable} also
build static and dynamic terms in tandem.  However, they
build a static term for every object term, even a bound variable,
so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remain untyped.  
The idea of generating both static and dynamic components alongside each
other seems to be in found in an early form in \citet[\S5]{consel-schism-93}. 
In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.

@@

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |static|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |static|, so again we need to change the type arguments to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.
The verbosity of these types is the only difficulty in defining a
replacement signature for |Symantics| which captures that of |CPST|
as well.

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.
The techniques we use to form such a family of interpreters
find their origins in \citearound{'s \emph{language
triplets}}\citet{Holst-AMIX}, though in an untyped setting.
\Citet{JonesNielson:94:AbstractInterpretation} also
prefigured this separation when they decomposed a denotational definition of an
untyped object language into a core semantics (which we call abstract syntax)
and multiple interpretations.
